Preliminary insights into the use of pottery and culinary pratices in southwestern China

### 2023-11-26 analysis/paper/paper.Rmd
revising based on comments

---
title: "Investigating Neolithic pottery use from Guijiabao in the Yanyuan Basin in southwestern China using organic residue analysis"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestrial animals across areas in southwestern China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region. It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and millets came even earlier. However, to what extent new crops were used by local peoples at that time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwestern China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C3 plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore the extent of the adoption of farming during the Neolithic period in southwestern China.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local people groups involved in the interactions [@Burmeister2000]. The Hengduan Mountain ranges and valleys in southwestern China are geographical areas that connect the Qinghai-Tibetan Plateau and the Yunnan–Guizhou Plateau and by which movement to these plateaus must occur. They play an important role in the cultural exchanges and spread of crops into southwestern China and even further to mainland Southeast Asia [@Ma2021; @Lin2022; @Hao2022; @Tong1986]. Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern Sichuan around 4700 BP and in western Yunnan before 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022; @Chen2015]. The Hengduan Mountain regions in the Neolithic period are therefore essential for understanding the processes of agricultural expansion. Recent archaeological evidence from the Yanyuan Basin in the middle Hengduan Mountains further reveals that millet and rice were present at roughly the same period as the grains found in Sichuan [@Huan2022; @Hao2022]. However, to what extent the two crops were incorporated into the local diet is still not fully understood.

To explore whether new crops or terrestrial animals became important food sources during the Neolithic period in the Yanyuan Basin, we investigated food residues within ceramic vessels from Guijiabao, a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a]. The archaeobotanical evidence at Guijiabao reveals that foxtail millet, broomcorn millet, and rice were present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Huan2022], and millet could have already been part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period. To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery? Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period? We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.

The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins [@Copley2005; @Whelton2021]. This is possible due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020]. Through our case study, we attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion.

@@

<!--the major finding at GJB-->
The common vessel forms unearthed at the Guijiabao site include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between the Guijiabao site and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal that cultivated crops were dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C4-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) has also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences [@Huan2022].

@@

## Samples
We selected ceramic fragments that were identified as typical from the Guijiabao and Gaoshan sites. The fragments date to the Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)), which were determined from known contexts based on radiocarbon dating. Ceramic fragments from the Guijiabao site were excavated from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that these ceramics were used in daily life [@Guijiabao2017a; @Hao2022]. This research also attempts to confirm that these pottery vessels were used as cooking/storage wares for food.

@@

## Methods
We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C~16:0~) and stearic (C~18:0~), which are found in most animals and plants, we identified their stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2004; @Craig2007; @Evershed1994].

For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C). We conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene prior to GC-MS.

@@

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector. Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness). The carrier gas was helium set at a constant flow rate of 1.3 ml/min. The temperature of the oven was set at 90 °C and held for 1 minute and then was increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode. The carrier gas was helium flowing at a rate of 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was set at 90°C and held for 1.3 minute, and then was raised to 320°C at a rate of 5°C/min. The final temperature was set for 22 minutes. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface. The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then was raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute.

@@

```{r highlight-compounds, cache=TRUE}
# highlight target compounds
highlight_common <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C14:0",
    retention == 23.6 ~ "C15:0",
    retention == 25.5 ~ "C16:0",
    retention == 27.3 ~ "C17:0",
    retention == 28.4 ~ "C18:1",
    retention == 29.1 ~ "C18:0",
    retention == 32.3 ~ "C20:0",
    retention == 35.4 ~ "C22:0")) %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
  filter(File > 100000)

@@

# highlight the compounds in the potsherd with high concentration
highlight_SYG2_2 <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C14:0",
    retention == 25.5 ~ "C16:0",
    retention == 27.3 ~ "C17:0",
    retention == 28.4 ~ "C18:1",
    retention == 29.1 ~ "C18:0",
    retention == 31.0 ~ "C19:0",
    retention == 32.4 ~ "C20:0",
    retention == 35.4 ~ "C22:0",
    retention == 36.8 ~ "C23:0",
    retention == 38.2 ~ "C24:0",
    retention == 40.9 ~ "C26:0",
    retention == 33.9 ~ "IS",
    retention == 41.3 ~ "IS",
    retention == 45.6 ~ "C30:0",
    retention == 47.8 ~ "C32:0")) %>%
  filter(name == "SYG-TN13-E22-2#2")
```

```{r chromatogram-four, cache= TRUE}
# plot the first four
chroma_four <-
  meth_all %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", 
                     "SYG-TN13-E23-3", "CDG-062")) %>%
  mutate(File = ifelse(File > 1000000, 1000000, File)) %>%
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  ggrepel::geom_text_repel(data = highlight_common, #ggrepel::geom_text_repel
            aes(label = FA, angle = 0),
            size = 3,
            nudge_y = 120000,
            show.legend = FALSE) +
  geom_text(data = highlight_IS,
            aes(label = FA),
            size = 3,
            nudge_y = 100000,
            show.legend = FALSE) +
  geom_segment(aes(x= 41.1, xend= 41.5, y= 1000000, yend= 1000000), size =0.3) +
  geom_segment(aes(x= 41.1, xend= 41.5, y= 1030000, yend= 1030000), size =0.3) +
  annotate("text", x = 41.3, y = 600000, label = "IS", size =3) +
  facet_wrap(~name, ncol = 1) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 1100000),
                     breaks = seq(0, 1100000, 200000)) +
  scale_x_continuous(limits = c(20, 45),
                     breaks = seq(20, 45, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(x = "retention time", y = "relative intensity") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the presence of lipid compounds were above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid), C~18:0~ (stearic acid), and monounsaturated fatty acid C~18:1~ (oleic acid) (Figure \@ref(fig:chromatograms-plots-four)). These fatty acids commonly exist in animal fat and plant oil. They are also generally degraded from larger molecules, such as triacylglycerols, due to oxidation and microbial degradation after long-term burial processes [@Dudd1998; @Eerkens2007]. High concentrations of medium-chain fatty acids (C~16:0~ and C~18:0~) with an even-over-odd carbon chain length predominance indicate that the organic residues in our potsherds were highly degraded during post-depositional processes. In addition, a number of potsherds had higher contents of C~18:1~ compared to C~18:0~ and C~16:0~, such as SYG-TN13-E23-3. A higher content of C~18:1~ and C~16:0~ with a lower content of C~18:0~ might indicate the origin of plant oils [@Mayyas2018; @Copley2005]. However, we may still need to examine their isotopic values to confirm their sources.

@@

quant_FA_tidy <-
  quant_FA %>%
  pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
               names_to = "FA", 
               values_to = "weight") %>% 
  filter(str_starts(FA, "ng") & !sample =="") %>% 
  mutate(FA = str_remove(FA, "ng "),
         weight = round(weight/`%reco`/`g extracted`/1000, 2)) %>% 
  mutate(weight = ifelse(is.na(weight)|weight == "0", "-", weight)) %>% 
  select(-number, -`%reco`, -`g extracted`) %>% 
  pivot_wider(names_from = sample, 
              values_from = weight) %>% 
  select("FA", "SYG-TN13-E22-1", "SYG-TN13-E22-2#1", "SYG-TN13-E22-2#2",
         "SYG-TN13-E23-3", "SYG-3", "CDG-062") %>% 
  filter(!FA == "C34:0" & !FA == "C21:0")

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites. 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds. 'IS' indicates internal standard used for quantification, including FA C21:0 and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

>>>>

---
title: "Preliminary insights into the use of pottery and culinary pratices in southwestern China"
author:
  - Li-Ying Wang:
      email: liyingw@asihp.net
      institute: [AcS]
      correspondence: true
  - Kuei-chen Lin:
      email: kueichenlin@asihp.net
      institute: [AcS]
      correspondence: false
  - Zhiqing Zhou:
      email: zhouzhiQ2000@163.com
      institute: [InsC]
      correspondence: false
institute:
  - AcS: Institute of History and Philology, Academia Sinica 
  - InsC: Chengdu Institute of Cultural Relics and Archaeology
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
    The spread of new crops or terrestrial animals across areas in southwestern China during the Neolithic period is one of the important archaeological issues to understand the process of adoption of new culinary practices in this region. It is widely accepted that rice was introduced into southwest China in 4700 BP at the latest and millets came even earlier. However, to what extent new crops were used by local peoples at that time remains unclear. The excavation of Guijiabao in the Yanyuan basin in southwestern China uncovered abundant materials with well-preserved stratigraphic records and has provided a unique opportunity to understand the local adoption of new crops. To further understand pottery use and food practices at Guijiabao, we investigated organic residues preserved in potsherds using lipid biomarker identification and compounds-specific stable isotope analysis. The results show signals of terrestrial animals in most potsherds and a C~3~ plant source in one potshard, during the Neolithic period. Our case study indicates that hunting might have been the primary means for obtaining subsistence, supplemented by rice/millet farming. This study also demonstrates an approach using organic residue analysis to explore the extent of the adoption of farming during the Neolithic period in southwestern China.

@@

# Introduction
<!-- intro to the topic and research aim-->
Movements of people between geographically distinct areas are usually accompanied by the exchange of goods, technology, or ideas that could have cultural or economic impacts on local populations involved in the interactions [@Burmeister2000]. In prehistoric southwest China, the Hengduan Mountain ranges have served as passages for people to move between the Qinghai-Tibetan Plateau in the north, the Yunnan–Guizhou Plateau in the south, and the Sichuan Basin in the east. The movements in the Hengduan Mountain ranges facilitated not only cultural exchange, but also the spread of new subsistence practices, such as farming and animal domestication [@Tong1986; @Hein2014; @Chen2015]. With more archaeological sites have been found recently along the Hengduan Mountain valleys, researchers are able to dive deeper into local cultural developments and diversity, the timing and routes of exchange network, and the spread of farming practices during Neolithic period across the Hengduan Mountain ranges and neighboring areas [@Hein2014; @Hein2016; @Ma2021; @Lin2022; @Hao2022; @Ma2021].

Despite a long-standing debate on the routes of rice/millet dispersal from northwest China and the middle Yangtze valley to southwestern China [@Guedes2013; @DAlpoim2018; @Liu2022], it is widely accepted that mixed farming was practiced in southwestern China as early as 4700 BP, where the adoption of farming in southwestern Sichuan is around 4700 BP and in western Yunnan is around 4600 BP [@Jiang2016; @YunnanMuseum1981; @Hao2022; @Zhang2010; @Dal2022]. Recent excavations at Guijiabao in the middle Hengduan Mountain ranges bridges the gap between southwestern Sichuan and Yunnan through the finding of millet and rice around 4700 BP [@Huan2022; @Hao2022]. However, it is still not fully understood the extent of the adoption of farming practices by the ancient people in the Hengduan Mountain ranges. Did crop cultivation become the essential part of local subsistence when they were introduced? What other resources were obtained in addition to those grains?

In this study, we focus on archaeological sites

aim to to explore the culinary practices through food residues within ceramic vessels from Guijiabao and Gaoshan sites

Also, the discussion of local subsistence mostly relies on paleobotanical evidence and zooarchaeological assemblages,  The preservation of animal or plant remains could be heavily impacted by environmental conditions.

to explore the culinary practices in the Yanyuan Basin through food residues within ceramic vessels from Guijiabao and Gaoshan sites. We analyzed small samples to test whether those pottery were used for cooking, and if we could detect new crops or species such as rice, millet, or pigs, as food sources when they spread into the Yanyuan Basin. Guijiabao is a long-term occupation settlement with complete temporal sequences from the Neolithic period to the historical period in Yanyuan that have been excavated so far [@Guijiabao2017a; @Guijiabao2020a]. The archaeobotanical evidence at Guijiabao reveals that foxtail millet, broomcorn millet, and rice were present in the Neolithic period around 5000 BP, earlier than previously thought [@Guijiabao2016; @Huan2022], and millet could have already been part of the local diet [@Lin2022]. Although some of these analyses suggest the presence of mixed millet/rice farming [@Huan2022], we still lack direct evidence of culinary practices during this period. To understand the extent of consuming new crops or terrestrial animals at Guijiabao, we ask the questions: What were the culinary practices that can be deduced by the pottery? Can we identify the new crops or terrestrial animals that were incorporated into the local food diets during the Neolithic period? We answer these questions by examining potsherds using organic residue analysis, which allows us to directly explore food cooked or processed in ceramic vessels.

The recent growing application of organic residue analysis demonstrates its usefulness in understanding ancient pottery use [@Kimpe2004; @Mayyas2022], human subsistence strategies, or culinary practices across a region over time [@Kwak2020; @Craig2011; @Suryanarayan2021; @Zhang2022; @Junno2020]. Examining organic residues, particularly lipids, enables archaeologists to investigate ancient food practices through lipid distributions and specific lipid compounds as biomarkers for tracing food origins; for example, millet can be identified through the biomarker miliacin [@Copley2005; @Whelton2021; @Heron2016]. This is possible for prehistoric ceramics due to the protection of clay matrices from microbiological degradation and water leaching, which allows lipid molecules to accumulate and be well preserved in the porous fabric of pottery for over a thousand years [@Evershed1993; @Evershed2008; @Roffet2017]. In this study, we detected and identified lipid residues absorbed into the ceramic matrix that can reveal the original contents cooked or stored in a vessel during its lifetime of use [@Dudd1999; @Miller2020]. Through our case study, we attempt to investigate culinary practices during a period of frequent cultural interactions that accompanied agricultural expansion.

@@

<!--the major finding at GJB-->
The common vessel forms unearthed at the Guijiabao site include jars, urns, bowls, and basins. The representative ceramic styles in the Neolithic period are grayish high-tempered fine ware and reddish brown coarse ware with decorations such as cord-marked and basket-impressed patterns [@Hao2022; @Guijiabao2017a]. Some ceramic decorations present northern styles, and some ceramics with incised geometric patterns and impressed dot patterns resemble those found in western Yunnan, indicative of relationships between the Guijiabao site and sites in Yunnan and the southern part of the Hengduan Mountains [@Guijiabao2017a]. Plant remains reveal that cultivated crops were dominated by foxtail millet (*Setaria italica*) and broomcorn millet (*Panicum miliaceum*) based on flotation samples collected from ash pits and stratigraphic layers that can be traced back to 5000 BP [@Guijiabao2016; @Hao2022]. This corresponds to a C~4~-based diet based on isotopic analysis of four human bones [@Lin2022]. In addition to millet, a smaller number of rice grains (*Oryza sativa*) has also been found in ash pits, where the earliest one dates to 4700 BP [@Hao2022]. A detailed phytolith analysis for soil samples dating to the Neolithic period (5000-3700 BP) further suggests the presence of rice phytoliths throughout the temporal sequences [@Huan2022].

@@

## Samples
As a preliminary research to accesses the feasibility of this approach, we selected ceramic fragments that were identified as typical from the Guijiabao and Gaoshan sites. The fragments date to the Neolithic (around 4500 BP) and historical (around 1000 BP) (Table \@ref(tab:sherds-table)), which were determined from known contexts based on radiocarbon dating. Ceramic fragments from the Guijiabao site were excavated from trench 13 excavated during the 2017 field season, along with one fragment from the 2016 season. The pottery samples are mostly gray base fragments with fine sand temper, of which a few have impressed decorations. Although their original forms could not be recovered, the evidence of chars on some fragments and their contexts in cultural layers suggest that these ceramics were used in daily life [@Guijiabao2017a; @Hao2022]. This research also attempts to confirm that these pottery vessels were used as cooking/storage wares for food.

@@

## Methods
We detected and identified lipid compounds using gas chromatography mass spectrometry (GC-MS) and quantified lipids using gas chromatography with flame-ionization detection (GC-FID). To distinguish the source of the common fatty acids, palmitic (C~16:0~) and stearic (C~18:0~), which are found in most animals and plants, we conducted compound-specific isotope analysis to identify stable carbon isotope characteristics using gas chromatography combustion isotope ratio mass spectrometry (GC-C-IRMS). The stable carbon isotope ratio for individual lipid compounds is a powerful indicator of the origins of fatty acids since different food categories, such as terrestrial animals, marine animals, and plants, have specific ranges of δ^13^C values [@Craig2007; @Evershed1994].

For sample processing, the surface of each potsherd was removed to prevent external contamination from environmental factors and handling during excavation or post-excavation analysis. Approximately 3.5 grams of powdered samples were extracted with a mixture of dichloromethane/methanol solution (9:1 v/v) using accelerated solvent extraction (ASE Dionex 350; at 100 °C) [@Mcclure2018]. To recover free fatty acids from bound lipids for a higher yield, we conducted saponification by adding 10 mL of 1M KOH in methanol and 1 mL water to the extracts and heating them in an oven at 70 °C overnight [@Kaluzna2016; @Craig2004]. Saponified extracts were acidified with 6M hydrochloric acid (HCl), and the acid fraction was extracted using hexane ($3\times10$ mL). A 5% aliquot of each sample was derivatized using 20 uL N,O-bis(trimethylsilyl)trifluoroacetamide (BSTFA) with 1% v/v trimethylchlorosilane (TMCS), and heated at 70 °C for one hour. After drying under a stream of nitrogen, the derivatized sample was dissolved in toluene prior to GC-MS.

@@

The GC-MS analysis was performed using an Agilent 7683 B series gas chromatograph attached to an Agilent 5975 inert mass selective detector (Agilent technologies, USA). Samples were injected in splitless mode at 320 °C, and compounds were separated with a VF-17ms column (60 m length, 0.25 mm i.d., 0.25 μm film thickness; J&W Scientific, USA). The carrier gas was helium set at a constant flow rate of 1.3 ml/min. The temperature of the oven was set at 90 °C and held for 1 minute and then was increased at a rate of 5 °C/min to 320 °C, where it was held for 23 minutes. The GC-FID analysis for the FAME samples was carried out using Agilent 6890N gas chromatograph equipped with a PTV inlet in splitless mode (Agilent technologies, USA). The carrier gas was helium flowing at a rate of 1.6 ml/min. The GC system was an Agilent 6890N equipped with a programmed temperature vaporization (PTV) inlet in splitless mode. The GC oven was set at 90°C and held for 1.3 minute, and then was raised to 320°C at a rate of 5°C/min. The final temperature was set for 22 minutes. The GC-C-IRMS analyses of FAMEs of individual fatty acids were conducted using a Thermo Delta V Plus mass spectrometer linked to a GC Trace 1310 through an Isolink Interface (Thermo Fisher Scientific, USA). The sample injection volume was 1 ul in toluene and was injected in splitless mode. Separation was carried out using the same model of column for GC-MS and the GC oven temperature was set at 90 °C for 1.5 minutes and then was raised at a rate of 5 °C/min to 320 °C, with an 18-minute hold. Helium was used as a gas carrier with a constant flow rate of 1.24 ml/minute.

@@

```{r highlight-compounds, cache=TRUE}
# highlight target compounds
highlight_common <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C[14:0]",
    retention == 23.6 ~ "C[15:0]",
    retention == 25.5 ~ "C[16:0]",
    retention == 27.3 ~ "C[17:0]",
    retention == 28.4 ~ "C[18:1]",
    retention == 29.1 ~ "C[18:0]",
    retention == 32.3 ~ "C[20:0]",
    retention == 35.4 ~ "C[22:0]")) %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", "SYG-TN13-E23-3", "CDG-062")) %>%
  filter(File > 100000)

@@

# highlight the compounds in the potsherd with high concentration
highlight_SYG2_2 <-
  meth_all %>%
  mutate(retention = round(Path, 1)) %>%
  group_by(sample, retention) %>%
  arrange(desc(File)) %>%
  distinct(retention, .keep_all = TRUE) %>%
  mutate(FA = case_when(
    retention == 21.6 ~ "C[14:0]",
    retention == 25.5 ~ "C[16:0]",
    retention == 27.3 ~ "C[17:0]",
    retention == 28.4 ~ "C[18:1]",
    retention == 29.1 ~ "C[18:0]",
    retention == 31.0 ~ "C[19:0]",
    retention == 32.4 ~ "C[20:0]",
    retention == 35.4 ~ "C[22:0]",
    retention == 36.8 ~ "C[23:0]",
    retention == 38.2 ~ "C[24:0]",
    retention == 40.9 ~ "C[26:0]",
    retention == 33.9 ~ "IS",
    retention == 41.3 ~ "IS",
    retention == 45.6 ~ "C[30:0]",
    retention == 47.8 ~ "C[32:0]")) %>%
  filter(name == "SYG-TN13-E22-2#2")
```

```{r chromatogram-four, cache= TRUE}
# plot the first four
chroma_four <-
  meth_all %>%
  filter(name %in% c("SYG-3", "SYG-TN13-E22-2#1", "SYG-TN13-E22-1", 
                     "SYG-TN13-E23-3", "CDG-062")) %>%
  mutate(File = ifelse(File > 1000000, 1000000, File)) %>%
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  ggrepel::geom_text_repel(data = highlight_common, #ggrepel::geom_text_repel
            aes(label = FA, angle = 0),
            size = 3,
            nudge_y = 120000,
            show.legend = FALSE,
            parse = TRUE) +
  geom_text(data = highlight_IS,
            aes(label = FA),
            size = 3,
            nudge_y = 100000,
            show.legend = FALSE) +
  geom_segment(aes(x= 41.1, xend= 41.5, y= 1000000, yend= 1000000), size =0.3) +
  geom_segment(aes(x= 41.1, xend= 41.5, y= 1030000, yend= 1030000), size =0.3) +
  annotate("text", x = 41.3, y = 600000, label = "IS", size =3) +
  facet_wrap(~name, ncol = 1) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 1100000),
                     breaks = seq(0, 1100000, 200000)) +
  scale_x_continuous(limits = c(20, 45),
                     breaks = seq(20, 45, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(x = "retention time", y = "relative intensity") +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

@@

The GC-MS analysis for organic residues extracted from potsherds found at the Guijiabao and Gaoshan sites revealed that the amount of lipid compounds ranges from 7 to 48.6 μg/g, above interpretable concentrations (5 μg/g) [@Evershed2008experimental], but were in relatively small amounts in the majority of the potsherds (Table \@ref(tab:FA-composition-table)). The lipid profiles show that the main compounds in most potsherds were saturated fatty acids with an even number of carbons, particularly C~16:0~ (palmitic acid), C~18:0~ (stearic acid) (Figure \@ref(fig:chromatograms-plots-four)). The fatty acids C~16:0~, C~18:0, and C~18:1~ commonly exist in animal fat and plant oil. Unsaturated fatty acid C~18:1~ was also found in all samples, but we are unable to distinguish the C~18:1~ types based on current method. In addition, a number of potsherds had higher contents of C~18:1~ compared to C~18:0~ and C~16:0~, such as SYG-TN13-E23-3.

@@

quant_FA_tidy <-
  quant_FA %>%
  rename("ng total" = "total FAs ug/g") %>%
  pivot_longer(!number&!sample&!`g extracted`&!`%reco`,
               names_to = "FA", 
               values_to = "weight") %>% 
  filter(str_starts(FA, "ng") & !sample =="") %>% 
  filter(!FA == "ng C34:0" & !FA == "ng C21:0") %>% 
  mutate(FA = str_remove(FA, "ng")) %>% 
  mutate(weight = ifelse(str_detect(FA, "C"), 
                         round(weight/`%reco`/`g extracted`/1000, 2), weight)) %>% 
  select(-number, -`%reco`, -`g extracted`) %>% 
  mutate(weight = round(weight, 2),
         weight = ifelse(is.na(weight), "-", weight),
         weight = ifelse(weight == 0, "-", weight)) %>% 
  pivot_wider(names_from = sample, 
              values_from = weight) %>% 
  select("FA", "SYG-TN13-E22-1", "SYG-TN13-E22-2#1", "SYG-TN13-E22-2#2",
         "SYG-TN13-E23-3", "SYG-3", "CDG-062") %>% 
  mutate(FA= str_replace(FA, "C(.*)", "C~\\1~"))

@@

```{r chromatograms-plots-four, echo=FALSE, fig.cap= "GC-MS chromatograms of fatty acids extracted from four potsherds from the Guijiabao and Gaoshan sites. 'Cn:x' refers to fatty acid with n carbon atoms and x double bonds. 'IS' indicates internal standard used for quantification, including FA C~21:0~ and 5-α Cholestane"}
knitr::include_graphics(here::here("analysis/figures/chromatograms_four.png"))
```

```{r chromatogram-specific, cache= TRUE}
# plot the specific one
chroma_specific <-
  meth_all %>%
  filter(name %in% c("SYG-TN13-E22-2#2")) %>% 
  ggplot(aes(Path, File)) +
  geom_line(size = 0.3) +
  ggrepel::geom_text_repel(data = highlight_SYG2_2, #ggrepel::geom_text_repel
            aes(label = FA),
            size = 3.5,
            nudge_y = 250000,
            show.legend = FALSE,
            parse = TRUE) +
  scale_y_continuous(labels = scales::comma_format(),
                     limits = c(0, 7500000),
                     breaks = seq(0, 7500000, 1000000)) +
  scale_x_continuous(limits = c(20, 50),
                     breaks = seq(20, 50, 5),
                     expand = c(0, 0.5)) + # don't log, many peaks and distort real counts
  labs(title = "SYG-TN13-E22-2#2", x = "retention time", y = "relative Intensity") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))
